Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-05-21 article/Old versions/MPS/PCADSC_article.tex
Changed manuscript according to reviewer comments (everything but component permutation issue) and styleguide.

@@

\renewcommand{\baselinestretch}{2}

@@

\begin{abstract}
Datasets are sometimes divided into distinct subsets, e.g. due to multi-center sampling, or to variations in instruments, questionnaire item ordering or mode of administration, and the data analyst then needs to assess whether a joint analysis is meaningful. The Principal Component Analysis-based Data Structure Comparisons (PCADSC) tools are three new non-parametric, visual diagnostic tools for investigating differences in structure for two subsets of a dataset through covariance matrix comparisons by use of principal component analysis. The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden and Bulgaria. The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of e.g. mean psychological well-being scores is not meaningful. However, when comparing Denmark and Sweden, very similar data structures, and thus comparable concepts of well-being, are found. Therefore, inter-country comparisons are warranted for these countries.
\end{abstract}

@@

Data comparability is a classical topic in statistics and a recurring one in psychometrics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically. This happens e.g. when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration or the order of survey questionnaire items are applied. All of these practices are essential parts of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of comparability among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability that help researchers identify potential problems or challenges early in the study design development are mostly absent. What is needed are procedures that compares differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. This paper suggests procedures for linear data structures Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.

While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection. Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well. However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies. This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for everyone to use. However, large scale publicly available datasets, such as the PISA (Programme for International Student Assessment) data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g. potential instrument-induced challenges in the datasets may not be available for these data analysts. For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.

The data structure comparison problem is however not limited to the context of new data sharing platforms. Even in single-population survey studies, a mixture of modes of administration is often used, e.g. mail and telephone, and while this can improve response rates, it can also be quite problematic \cite{Brambilla1987,McHorney1994}, and differences in response behavior can lead to biased results. Powers, Mishra and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle. Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016}

The comparability issue in these examples can be summarized as follows: Assume that we have two datasets with the same variables, but different observations (often represented as a single dataset with a subset-inducing variable) and that we wish to compare them without specifying a model, or even a variable of interest. The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be dealt with in later statistical modelling.

We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.

This manuscript describes the procedures, including a brief introduction to principal component analysis (PCA), and presents a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is completely ignored e.g. when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.

@@

PCADSC compares the covariance matrices of the two subsets of a dataset. If all variables in the datasets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing these entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.

At this point a general remark about standardization is in place. PCA deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will always be deemed highly influential, no matter the structure of the data. Therefore, the variables are typically standardized prior to performing PCA. Note that the covariance matrix of the standardized variables is the same as the correlation matrix of the original variables, so post-standardization PCA simply corresponds to performing data structure comparisons of the correlation matrices rather than the covariance matrices. The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation. In this paper we will adopt this standardization policy, and in particular, we will allow for differences between scalings of the original, raw variables in the two data subsets.

\subsection*{Using PCA for data structure comparisons}

PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful and interesting objects to compare across such different datasets. We consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ and let $\bar{x} = {(\bar{x}_1, ..., \bar{x}_d)}^\top = \frac{1}{n} \sum_{i=1}^n x_n$ and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) {(x_i-\bar{x})}^\top \in \RR^{d \times d}$ denote their averages and the empirical covariance matrix of the full data matrix, respectively. A brief introdution to CFA is included in the supplementary material.

Usually, in PCA, the main interest lies in the \textit{scores}, i.e. the coordinates $\eta_j^\top (x_i - \bar{x})$ of the projections observations onto the loadings, but the variation components and the accompanying loadings also describe aspects of the data. If two different datasets with the same variables, but different samples of observations, have similar loadings, then the variables can be argued to measure the same underlying quantities in both datasets. Similar loadings can occur in situations where the two sets of scores are different, e.g.\ if the two datasets come from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variable interplay differs in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the same $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. %
Furthermore, let $X$ be the combined dataset consisting of all $n=n_1+n_2$ observations. For these %two
three datasets, we complete the following steps:
\begin{enumerate}
\item Within each dataset, standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ and $\tilde{x}_j \in \RR^d$ for $j=1,\dotsc,n$ be the standardized datasets.
\item Form the principal component analyses
\begin{align*}
S_i &= \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$,} \\
S &= \frac{1}{n-1} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top,
\end{align*}
thereby obtaining loadings $\eta_{1k}, \eta_{2k}, \eta_k$ and variance components $\lambda_{1k}, \lambda_{2k}, \lambda_k$ for $k=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences. These three plots are:
\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

Note that the standardization implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equal 1, and thus also that
\begin{equation*}
\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d.
\end{equation*}
This identity will simplify some expressions below.

While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}

In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models (S1). Further details about data simulation is also available in the supplementary material. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\begin{figure}[h]
\caption{[FIGURE 1 ABOUT HERE]. Graphical representations of models used for simulating example data. Square nodes represent observed variables, while rounded nodes represent latent variables. Arrows are used to illustrate causal links.}
\label{fig:simGraphs}
\end{figure}

\subsubsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1,\lambda_{11}-\lambda_{21}), \;
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}.

\begin{figure}[h]
\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
\label{plot.simCE} 
\end{figure}

In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected). This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data. In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups. In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} {\bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)}^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

If two countries differ e.g. in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material (S2). We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

@@

\subsection*{Comparing Denmark and Bulgaria}
Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales. The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures. This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.

\begin{figure}[h]
\caption{[FIGURE 5 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
\label{plotBG.cehair}
\end{figure}

>>>>

\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes, positioning, calc}

\usepackage{float}

@@

\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\renewcommand{\baselinestretch}{2}

@@

\begin{abstract}
Datasets are sometimes divided into distinct subsets, e.g. due to multi-center sampling, or to variations in instruments, questionnaire item ordering or mode of administration, and the data analyst then needs to assess whether a joint analysis is meaningful. The Principal Component Analysis-based Data Structure Comparisons (PCADSC) tools are three new non-parametric, visual diagnostic tools for investigating differences in structure for two subsets of a dataset through covariance matrix comparisons by use of principal component analysis. The PCADCS tools are demonstrated in a data example using European Social Survey data on psychological well-being in three countries, Denmark, Sweden, and Bulgaria. The data structures are found to be different in Denmark and Bulgaria, and thus a comparison of for example mean psychological well-being scores is not meaningful. However, when comparing Denmark and Sweden, very similar data structures, and thus comparable concepts of well-being, are found. Therefore, inter-country comparisons are warranted for these countries.
\end{abstract}

@@

Data comparability is a classical topic in statistics and a recurring one in psychometrics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically. This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration, or the order of survey questionnaire items are applied. While a mixture of modes of administration, for example mail and telephone, can improve response rates in surveys, it can also be quite problematic by inducing differences in response behavior that may lead to biased results \cite{Brambilla1987,McHorney1994}. Powers, Mishra, and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle. Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016} But mixing several data collection methods is an essential part of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of comparability among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

The comparability issue in these examples can be summarized as follows: Assume that we have two datasets with the same variables, but different observations, often represented as a single dataset with a subset-inducing variable, and that we wish to compare them without specifying a model, or even a variable of interest. The central question is then whether the two datasets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity that must be dealt with in later statistical modelling.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model. But this places the effort of assessing data comparability very late in the data analysis process and it makes the comparability assessment ad-hoc, as it essentially relies on the appropriateness of the modeling choices, which again depends on the structure of the data. The use of (parametric) models does thus not constitute a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. However, useful tools for initial, exploratory investigations of data comparability %that help researchers identify potential problems or challenges early in the study design development 
are mostly absent. What is needed are procedures that compare differences in data structures in two subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables.  Simple methods like variable-by-variable tests of distributional differences suffer from the drawback that they only address marginal differences and ignore the interplay between variables, but entry-by-entry comparisons of two empirical correlation matrices quickly become unmanageable as the number of variables increase.

%While the issue is not in any way new, its relative importance is increasing due to new developments in data availability and data collection. Classical statistical methods are generally aimed at analyzing data from designed experiments and historically, statistical analyses have been conducted by researchers who knew the design and the source of the dataset well. However, the origin stories of datasets have changed over time and today, a lot of data are accumulated without a specific purpose in mind, as data collection and sharing has become easier and more affordable through the development of new technologies. This new era of data sharing facilitates a tremendous amount of new research, as data on all sorts of topics are now readily available online for everyone to use. However, large scale publicly available datasets, such as the PISA (Programme for International Student Assessment) data and European Social Survey (ESS) data, are often used by data analysts that are far removed from the data producers and therefore, problem-specific recommendations about e.g. potential instrument-induced challenges in the datasets may not be available for these data analysts. For this reason, a thorough investigation of data structure comparability becomes a crucial step in any meaningful data analysis when using such open-source data.

We propose three visual tools for data structure comparisons, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The use of PCA implies that we can only compare datasets with numerical variables and that we focus on the linear aspect of the  data structures. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}. Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is ignored e.g. when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.

@@

PCADSC compares the covariance matrices of two subsets of a dataset. If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets. We refer to Appendix S1 for a brief introduction to PCA.

We assume that the variables have been standardized before PCA is conducted, which is equivalent to working with the empirical correlation matrices rather than the empirical covariance matrices. We also assume that all considered variables have a numerical interpretation, for example by being continuous or ordinally categorical. Moreover, we use the following notation: 
$\mathbf{X} \in \RR^{n_x \times d} $ and $\mathbf{Y} \in \RR^{n_y \times d} $ are datasets containing the same number of variables, $d$, but possibly different numbers of observations, $n_x$ and $n_y$. We use $X_j$ to refer to the $j$th variables of $\mathbf{X}$ and $x_{ij}$ to refer to the $i$th observation  within that variable, while $x_{i \cdot}$ is the full $i$th observation row. Note that $X_j \in \RR^{n_x}$, $x_{ij} \in \RR$, and $x_{i \cdot} \in \RR^d$. We let $\bar{X} = (\bar{X}_1, ... , \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X}) - (x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.

\subsection*{Using PCA for data structure comparisons}
%PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and eigenvalues are meaningful and interesting objects to compare across such different datasets.

The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{n_x + n_y \times d}$$
 be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
\item Form the principal component analyses
\begin{align*}
S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
\end{align*}
thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
\begin{equation*}
\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
\end{equation*}
This identity will simplify some expressions below. Using a greedy interpretation of PCA, which is often overlooked in the literature, we can note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences, namely the \textit{cumulative eigenvalue (CE) plot}, the \textit{angle plot}, and the \textit{chroma plot}. These plots are dsecribed in turn below. For the deepest understanding of the data structure differences in two datasets, we suggest using all of the three plots in the same order as they are presented. While we describe the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle}, and \ref{plot.simChroma} for the CE-, angle-, and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}

In \textit{Model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix S2. In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

%\begin{figure}[h]
%\caption{[FIGURE 1 ABOUT HERE]. Graphical representations of models used for simulating example data. Square nodes represent observed variables, while %rounded nodes represent latent variables. Arrows are used to illustrate causal links.}
%\label{fig:simGraphs}
%\end{figure}

\subsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
\end{align*}
Due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are. Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does. By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more infuential components drive the visual impression.

%\begin{figure}[h]
%\caption{[FIGURE 2 ABOUT HERE] CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B %(where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.}
%\label{plot.simCE} 
%\end{figure}

Two examples of cumulative eigenvalue plots are available in Figure \ref{plot.simCE}. In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected). This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data. In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups. In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k^z + \lambda_{k+1}^z}{2} {\bigg( \sum_{j=1}^k \lambda_{j}^x - \sum_{j=1}^k \lambda_{j}^y \bigg)}^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_x$ and $n_y$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing}, and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in Appendix S3. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

@@

\subsection*{Comparing Denmark and Bulgaria}
Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales. The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures. This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero. The CE curve generally lies above zero, suggesting that the Bulgarian dataset has larger eigenvalues for all but the last component, thereby explaining more variance at lower dimensionalities.

%\begin{figure}[h]
%\caption{[FIGURE 6 ABOUT HERE] CE plots (top panel) and angle plots (bottom panel) comparing Danish and Bulgarian data on psychological well-being. %The CE plot is annotated with the $p$-values of the Kolmogorov-Smirnov and the Cram\'er-von Mises tests of the hypothesis of no difference in data structures. In the angle plot, the blue arrows show the principal components of the Bulgarian dataset decomposed in the coordinate system of the principal components of the Danish dataset, while the red arrows illustrate the reverse.}
%\label{plotBG.cehair}
%\end{figure}
